---
output: html_document
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 80
---

# Assesing uncertanties related to satellite remote sensing indices to estimate Gross Primary Production

## Introduction

Vegetation Gross Primary Production (GPP) is the total amount of carbon fixation
by plants through photosynthesis [@badgley_terrestrial_2019]. Quantifying GPP is
essential for understanding land-atmosphere carbon exchange
[@kohler_global_2018], ecosystem function, and ecosystem responses to climate
change [@guan_comparison_2022; @brown_fiducial_2021; @myneni_relationship_1994].
However, terrestrial GPP cannot be directly estimated due to the contribution of
respiration to land carbon fluxes [@anav_spatiotemporal_2015]. Instead, GPP can
be inferred in a nondestructive manner by the net carbon exchange measurements 
at the ecosystem level, or at broarder scales using models that incorporate 
various assumptions and limitations [@reichstein_separation_2005; @jung_towards_2009]. 
One widely used approach for approximating and quantifying GPP involves employing
data-driven models based on spectral vegetation indices (VIs) [@ryu_what_2019] 
However, GPP estimation using spectral VIs remain with high uncertainties, 
prompting the need for validation efforts are required to identify approaches to
improve accuracy. [@anav_spatiotemporal_2015; @brown_fiducial_2021]

GPP estimations can be grouped into two broad categories: Eddy Covariance (EC) 
techniques, and satellite data-driven methods [@guan_comparison_2022; @xie_assessments_2020].
EC is the primary in-situ non-destructive method for measuring terrestrial
fluxes, and specifically for quantifying the exchange of CO2 between land
and the atmosphere, using advanced field instrumentation [@baldocchi_how_2020; @badgley_terrestrial_2019; @ryu_what_2019; @tramontana_predicting_2016]. However,
EC measurements come with certain limitations, such as their relatively low 
spatial resolution, typically less than \< 1km2 , which constrains the accuracy
of estimating ecosystem carbon and water fluxes at regional and global scales. [@badgley_canopy_2017].

Additionally, it's important to note that EC techniques directly measure Net
Ecosystem Exchange (NEE), not GPP. Subsequently GPP must be estimated by both
subtracting respiration using models and ancillary measurements while accounting
for removal or deposition of carbon stocks due to natural or anthropogenic
transport processes such as water flow, fires or harvest [@beer_terrestrial_2010; @reichstein_separation_2005].

The second category of methods to estimate GPP is the satellite data-driven
models. Because of their dependency on earth observation platforms, they are not
spatially constrained but can have more uncertainties than the EC techniques
[@ryu_what_2019; @wang_diagnosing_2011]. Satellite data-driven models can be
classified into process-based models, Light Use Efficiency models (LUE), and
Vegetation Index models [@xie_assessments_2020].

Process-based models integrate climate, canopy and soils information derived
from multiple sources, including satellite EO, into biophysical models of
carbon, water, energy, and nutrient cycles with varying levels of 
detail [@running_general_1988; @harris_global_2021]. While these models can be 
scaled globally [@beer_terrestrial_2010; @jung_towards_2009] they require many 
parameters that may not be readily available in changing landscapes or for fine
scale studies.

LUE models are based on the concept of radiation conversion efficiency, and take
into consideration ecological processes [@liu1997process; @heinsch_evaluation_2006].
This efficiency signifies the amount of carbon a specific vegetation type can 
fix per unit of solar radiation [@monteith_solar_1972]. Initially used for NPP
estimation [@prince_model_1991], LUE models were subsequently adapted for GPP 
and respiration calculations [@goetz_mapping_1999]. These models explicitly 
account for the impact of environmental stress on plant physiological responses.
The imposition of environmental stressors may lead to a reduction in the rates of 
daily carbon assimilation, thereby diminishing overall efficiency. [@prince_model_1991; @running_general_1988]. 

The GPP product from the Moderate Resolution Imaging Spectroradiometer (MODIS) 
employs an algorithm based on the radiation conversion efficiency concept. This
algorithm establishes a connection between absorbed photosynthetically active 
radiation (APAR) and the LUE term [@heinsch_evaluation_2006] as shown in @eq-gpp.

$$
\small
GPP = PAR \times fAPAR \times LUE
$$ {#eq-gpp}

Where `PAR` is the incident photosynthetically active radiation [@monteith2013principles] 
and `fAPAR` is the fraction of the PAR that is effectively absorbed by 
plants [@gcos200]. The `LUE` term depends on vegetation type but also 
physiological conditions driven by water availability, temperature stress, and
vapour pressure deficit [@goetz_mapping_1999; @running_general_1988]. Obtaining
these variables for every vegetation type on Earth can be challenging, 
introducing assumptions that amplify uncertainties [@goetz_mapping_1999].
Nevertheless, under unstressed conditions, LUE remains constant for a given 
vegetation type, requiring only PAR and fAPAR to assess primary 
productivity [@running_continuous_2004].

Satellite VIs have been used to apply @eq-gpp from regional to global extents 
[@running_continuous_2004]. VIs are a summary of non-linear functions of surface
bi-directional reflectance spectra [@myneni_relationship_1994] derived from 
optical sensors that are combined with climate variables to calculate GPP 
[@wu_predicting_2011]. This is usually done with some form of regression and 
physical methods that associate interactions between vegetation and incoming 
radiation [@fernandez-martinez_monitoring_2019]

VI's have been used to provide inputs to @eq-gpp related to fAPAR [@sellers_global_1994]
or to directly estimate GPP such as normalized difference vegetation index
(NDVI), enhanced vegetation index (EVI), Near-Infrared Reflectance Index (NIRv)
and Chlorophyll/Carotenoid Index (CCI) among others. [@balzarolo_influence_2019;
@rahman_potential_2005; @rahman_potential_2005; @xie_assessments_2020;
@badgley_terrestrial_2019; @zhang_potential_2020; @sellers_global_1994] These
VIs are based on a spectral reflectance ratio between the red and near-infrared
regions of the electromagnetic spectrum [@glenn_relationship_2008] which tracks
an integrated impact of fraction of photosynthetically active radiation (fAPAR)
and LUE on productivity.[@myneni_relationship_1994]

Nonetheless, even when estimating GPP from satellite remote sensing methods is
promising to overcome limitations as temporal and spatial scale, there are some
challenges involved in the process. The first constraint is the use of images
needed to address the problem of mixed-pixels, which requires determining the
fraction of the vegetated surface and its attributable signal
[@badgley_canopy_2017]. A second challenge is that photosynthesis regulation can
occur with no major changes in canopy structure or leaf pigments that can
undergo without being detected with reflectance data
[@pabon-moreno_potential_2022; @pierrat_diurnal_2022], and this could imply that
temporal aggregation may be critical for VI models.

The accuracy of GPP estimation through the use of VIs can be affected due to
their limitation to predict spatial, temporal, interannual variability and
trends. [@fernandez-martinez_monitoring_2019]. An index such as NDVI is good for
detecting changes in seasonal variability, but it becomes saturated with high
biomass conditions [@badgley_canopy_2017]. Other indices such as EVI can
overcome this problem and be better suited for predicting GPP in large biomass
forests [@badgley_canopy_2017] but it has been evaluated in a narrow range of
ecosystems and needs inputs of start and end dates of the growing seasons which
can increase uncertainties [@shi_assessing_2017].

To overcome these problems, other indices have been formulated in recent years.
One of these VIs is near-infrared reflectance index (NIRv)
[@badgley_terrestrial_2019] NIRv is defined as the fraction of reflected NIR
light that originates from vegetation. NIRv was originally proposed as a
replacement for fPAR in LUE models in that the NIR and PAR reflectance of
vegetation are correlated and the scaling by NDVI corrects for soil
contributions in the signal [@badgley_terrestrial_2019] NIRv has been shown to
have a stronger correlation to GPP at flux towers and on a regional basis than
fAPAR notwithstanding the fact that GPP is directly related to fAPAR. NIRv has
been extended to include a weighting with PAR [@dechant_nirvp_2022] and
replacing the NIR reflectance with NIRv radiance [@wu_radiance-based_2020]. Both
of these approaches have been shown to have even stronger correlations with GPP
than NIRv as could be expected since they either directly or indirectly weight
the NIRv with PAR.

The stronger correlation between the NIR indices to GPP versus indices based on
fAPAR seemingly contradicts the hypotheses in the LUE model that GPP should be
linearly related to fAPAR. There are two explanations: i. Many of the reported
comparative studies use fAPAR based on VIs and not APAR. We hypothesize that the
NIR indices are simply better estimators of APAR than these VI approximations
due to lower measurement error for the NIR indices or a stronger physical
relationship between them and APAR versus the historical VIs. ii. The strength
of correlation between APAR and GPP depends on LUE having a linear relationship
to observed APAR. While this may hold in some circumstances (e.g. early seasonal
measurements for vegetation such as crops where leaf chlorophyll concentration
increases during the growth phase) it is not the case in general and definitely
during stress conditions [@monteith_solar_1972].

Despite these efforts, relying solely on these remote satellite VIs presents a
challenge. The assumptions accompanying VI models suggest the importance of
systematically quantify their predictive capacity for GPP across various growing
seasons and locations. Therefore, there is a demand to make efforts to validate
and quantify their accuracy [@brown_fiducial_2021]. In-situ eddy-covariance flux
measurements represent a solution given that they represent site-level
observations. With the calculation of the flux footprint, the area from the flux
tower can serve as a reference point in which satellite images can be linked to
a specific set of pixels and calculate the accuracy of the estimations provided
by the indices [@chu_representativeness_2021].

As such, the main objective of my M.Sc. thesis is the analysis of uncertainty
associated with the VI/LUE models since they offer the potential for GPP
estimation without requiring knowledge regarding canopy and soil properties
other than measurements of reflected light and possibly a correction for soil
contribution. We aim to quantify the uncertainty from some of the most used
satellite vegetation indices with in-situ GPP data for multiple growing seasons
from the Bartlett Experimental Forest (USA), the Borden Forest Research Station
flux-site (Canada), and the University of Michigan Biological Station (USA).

\newpage

## Methods

```{r libraries and sources}
#| echo: false
#| message: false
#| warning: false

# Libraries
library(ggplot2)
library(cowplot)
library(lubridate)
library(purrr)
library(broom)
library(gt)
library(tidymodels)
library(broom)
library(usemodels)
library(vip)
library(h2o)
library(mgcv) 

# Source files
# Source the objects created for the complete GPP trends.
# This file will source the code and load objects to memory.
source("scripts/trend_plots.R")

# Source the objects created for the complete GPP trends
source("scripts/models_data_preparation.R")

# Source file with functions to plot rf predictions
source("R/plot_exploratory.R")
```

### Eddy Covariance sites

We used three deciduous broadleaf forest sites located in the northern
hemisphere with eddy covariance (EC) data collected by Ameriflux. For each site,
we used the daily, weekly, and monthly GPP values (GPP_DT_VUT_REF variable)
estimated using the ONEFlux workflow [@pastorello2020fluxnet2015]. The ONEFlux
processing does the estimation of the CO~2~ fluxes into GPP and Ecosystem
Respiration (RECO) from Net Ecosystem Exchange (NEE) through two methods known
as daytime and nighttime. Here we selected the daytime method (DT) which uses
daytime and nighttime to parameterize a model with two components: one based on
light response curve and vapour pressure deficit and a second one using a
respiration-temperature relationship to estimate RECO which in turn is used to
obtain the difference with NEE and provide GPP. [@pastorello2020fluxnet2015].

@fig-gpp_trends displays the GPP trends for the University of Michigan
Biological Station, Bartlett experimental forest, and the Borden Forest Research
Station. Additional details regarding the characteristics of these datasets can
be found in Table @tbl-oneflux_datasets.

| Site     | Data range available | Dataset name                                                | Reference                  |
|--------------------|---------------------|---------------------|--------------------|
| Bartlett | Jan 2015 to Dec 2017 | US-Bar: Barlett Experimental Forest (version: beta-3)       | [@staebler2019ameriflux]   |
| Borden   | Jan 2015 to Jan 2022 | CA-Cbo: Ontario - Mixed Deciduous, Borden Forest Site       | [@richardson2016ameriflux] |
| Michigan | Jan 2015 to Jan 2018 | US-UMB: Univ. of Mich. Biological Station (version: beta-4) | [@gough2016ameriflux]      |

: ONEFlux sites datasets description {#tbl-oneflux_datasets}

The Bartlett experimental forest is located in New Hampshire, USA (44°06′N,
71°3′W). This site is characterized by a forest with an average canopy height
ranging from 20 to 22 meters with a mean annual temperature of 6C. Despite
events such as a hurricane in 1938 and small scale forest management, the forest
mean stand age is around 120-125 years. [@ouimette_carbon_2018]

The second flux site is Borden Forest Research Station located in Ontario
(44°19′N, 79°56′W), Canada. This is one of the largest patches of forest in
Southern Ontario which has been collecting EC data since 1996
[@rogers_response_2020]. This site has a forest cover of over 60% with a height
that exceeds 2 m. It's a deciduous broadleaf natural re-growth forest dominated
by woody vegetation. [@lee_long-term_1999].

The third site is the University of Michigan Biological Station which is located
in northern Michigan, USA (45°350 N 84°430 W). The site have a forest with
different succesional stages and a mean height of 2m. Mean annual temperature is
around 5.5°C. [@gough_disturbanceaccelerated_2021]

```{r}
#| label: fig-gpp_trends
#| fig-cap: "GPP trends from the study sites on a daily (a), weekly (b), and monthly (c) basis for the University of Michigan Biological Station, Bartlett experimental forest, and the Borden Forest Research Station"
#| fig-width: 11
#| fig-height: 12
#| echo: false
#| message: false
#| warning: false
gpp_trends
```

<!-- ```{r} -->

<!-- #| label: fig-michigan_gpp_trends -->

<!-- #| fig-cap: "GPP trends for Michigan" -->

<!-- #| fig-width: 7 -->

<!-- #| fig-height: 9 -->

<!-- #| echo: false -->

<!-- #| message: false -->

<!-- #| warning: false -->

<!-- michigan_gpp_trends -->

<!-- ``` -->

<!-- ```{r} -->

<!-- #| label: fig-bartlett_gpp_trends -->

<!-- #| fig-cap: "GPP trends for Bartlett" -->

<!-- #| fig-width: 7 -->

<!-- #| fig-height: 9 -->

<!-- #| echo: false -->

<!-- #| message: false -->

<!-- #| warning: false -->

<!-- bartlett_gpp_trends -->

<!-- ``` -->

<!-- ```{r} -->

<!-- #| label: fig-borden_gpp_trends -->

<!-- #| fig-cap: "GPP trends for Borden" -->

<!-- #| fig-width: 7 -->

<!-- #| fig-height: 9 -->

<!-- #| echo: false -->

<!-- #| message: false -->

<!-- #| warning: false -->

<!-- borden_gpp_trends -->

<!-- ``` -->

<!-- The GPP for each of the sites can -->

<!-- -   include description of the ONEFlux processing (gap filling, data quality, -->

<!--     gpp estimation, gpp uncertainty with DT and NT) -->

### Satellite imagery

We used Google Earth Engine (GEE) to retrieve data from the Terra Moderate
Resolution Imaging Spectroradiometer (MODIS), specifically the collection
MOD09GA Version 6.1 product. A square polygon with an area of 3km surrounding
the EC tower was defined for each site, and the complete data pixel values
within this polygon was extracted for analysis.

The MODIS contains the surface spectral reflectance from bands to 1 through 7
with a spatial resolution of 500m, with corrections for atmospheric conditions
such as aerosols, gasses and Rayleigh scattering [@vermote2021modis]. Bands used
to derived the vegetation indices are shown in @tbl-MODIS_500_indices_bands

| **Name**    | **Description** | **Resolution** | **Wavelength** |
|-------------|-----------------|----------------|----------------|
| sur_refl_01 | Red             | 500 meters     | 620-670nm      |
| sur_refl_02 | NIR             | 500 meters     | 841-876nm      |
| sur_refl_03 | Blue            | 500 meters     | 459-479nm      |
| sur_refl_04 | Green           | 500 meters     | 545-565nm      |

: MODIS (MOD09GA.061 product) bands used to calculate the VIs
{#tbl-MODIS_500_indices_bands}

<!-- | Data                  | Spatial resolution (m) | GEE image collection         | -->

<!-- |-----------------------|------------------------|------------------------------| -->

<!-- | MODIS reflectance     | 500                    | MODIS/006/MOD09GA            | -->

<!-- | MODIS reflectance     | 250                    | MODIS/061/MOD09A1            | -->

<!-- | Harmonized Sentinel-2 | 20 - 60 - 10           | COPERNICUS/ S2_SR_HARMONIZED | -->

<!-- : Satellite images datasets {#tbl-satellite_images_datasets} -->

The data processing involved three main steps: selecting high-quality pixels,
scaling band values, and calculating vegetation indices. MODIS product data have
4 bit-encoded variables which provided information about the observation
quality. From those variables, only the `state_1km` and `qc_500m` variables were
used along with each of the bands bits quality indicators, as `q_scan` was not
informative and `g_flags` had the same value for all observations. The
bit-encoded variables were transformed into categorical strings, and only the
categories indicating the best quality were selected to filter the pixels
(@fig-quality_pixels). The specific bit strings selected for `state_1km` are
shown in @tbl-state_1km_bitstrings and for `qc_500m` in
@tbl-qc_scan_bit_strings.

In the second stage of the data processing, after selecting the highest quality
pixels, all the bands values were scaled. For each observation, regardless of
the specific band, a scaling factor of `0.0001` was applied. This scaling factor
was used to adjust the magnitude of the values and ensure consistency in the
data.

According to the documentation, any scaled value that fell outside the range of
`0` to `1` was considered a fill value or uncorrected Level 1B data and was
subsequently discarded. These values were deemed unreliable or lacking
meaningful information for the analysis. The complete selected high quality
pixels for the complete period per each of the sites is shown in
@fig-complete_quality_pixels

Once all the band values were scaled within the valid range, the vegetation
indices such as `NDVI` (@eq-ndvi), `NIRv` (@eq-nirv), `EVI` (@eq-evi), and `CCI`
(@eq-cci) were calculated and then matched with the corresponding date in the
flux datasets. The resulting values used for the analysis are shown in
@fig-quality_pixels

<!-- However, since the -->

<!-- MODIS product with a spatial resolution of 250m (`MODIS/061/MOD09A1`) does not -->

<!-- include the blue band EVI calculations were not performed for this dataset. -->

```{=tex}
\begingroup
\fontsize{10pt}{10pt}\selectfont
```
$$
\small
NDVI = \frac{NIR - Red}{NIR + Red}
$$ {#eq-ndvi}

$$
NIRv = NIR\times\frac{NIR - Red}{NIR + Red}
$$ {#eq-nirv}

$$
EVI = 2.5\times\frac{\mathrm{NIR} - \mathrm{Red}}{%
          (\mathrm{NIR} + 6\times \mathrm{Red}
          - 7.5\times \mathrm{Blue} + 1)}\\
$$ {#eq-evi}

<!-- $$ -->

<!-- kNDVI = tanh(\mathrm{NDVI}^2) -->

<!-- $$ {#eq-kndvi} -->

$$
CCI = \frac{Green - Red}{Green + Red}
$$ {#eq-cci} \endgroup

<!-- For Harmonized Sentinel-2 data the data processing consisted on using the -->

<!-- `msk_cldprb` variable to filter out pixels according to the probabilities of -->

<!-- containing a cloud or the `msk_snwprb` variable for presence of snow -->

<!-- probability. Once the pixels without any probability of having clouds or snow -->

<!-- were filtered from the rest, we scaled all the reflectance bands observations by -->

<!-- a factor of `0.0001`. With the values scaled we proceed to calculate the -->

<!-- vegetation indices `NDVI` (@eq-ndvis), `NIRv` (@eq-nirvs), and `EVI` (@eq-evis). -->

<!-- The bands used to calculate these indices are in -->

<!-- @tbl-harmonized_s2_indices_bands -->

<!-- $$ -->

<!-- NDVI = \frac{b8 - b4}{b8 + b4} -->

<!-- $$ {#eq-ndvis} -->

<!-- $$ -->

<!-- NIRv = b8\times\frac{b8 - b4}{b8 + b4} -->

<!-- $$ {#eq-nirvs} -->

<!-- $$ -->

<!-- EVI = 2.5\times\frac{\mathrm{b8} - \mathrm{b4}}{% -->

<!--           (\mathrm{b8} + 6\times \mathrm{b4} -->

<!--           - 7.5\times \mathrm{b2} + 1)}\\ -->

<!-- $$ {#eq-evis} -->

```{r}
#| label: fig-complete_quality_pixels
#| fig-cap: "Total number of observations (pixels) from MODIS classified as high quality (used in the analysis) or other quality (filtered out from the analysis)"
#| echo: false
#| message: false
#| warning: false
source("scripts/quality_observations.R")

all %>%
  filter(quality == "high") %>%
  mutate(year_mon = zoo::as.yearmon(date)) %>%
  ggplot(aes(x = year_mon, fill = site)) +
  geom_bar(position = "stack") +
  scale_fill_viridis_d(begin = 0.2, end = 0.8) +
  labs(x = "Date",
       y = "Total observations (pixels)",
       fill = "Site") +
  theme_bw(base_size = 12)
```

### Data Preparation

In the study, three datasets were prepared for each site: a daily dataset, a
weekly dataset, and a monthly dataset. These datasets were generated from the
satellite imagery data with the selected high quality pixels and the
ONEFluxprocess data in order to capture variations in vegetation indices (VI),
band values, and GPP over different time scales.

The daily dataset included the VI values, band values only from the high quality
pixels, and GPP measurements derived from the ONEFlux process collected on a
daily basis. This dataset provided a high-resolution representation of the
variables, allowing for a detailed analysis of their daily fluctuations.

The weekly and monthly datasets were derived from the corresponding daily
dataset. These datasets contained summarized values of the VIs and band values,
aggregated over the weekly and monthly time frames, respectively. The
aggregation process involved calculating an average for the VIs and band values
within each week or month.

For the GPP values in the weekly and monthly datasets, rather than summarizing
the daily GPP values, the GPP measurements for the weekly and monthly time
frames were obtained directly from the ONEFlux process, which provided a
reliable estimation of GPP for these longer time intervals.

By creating these three datasets (daily, weekly, and monthly), the study allowed
for a comprehensive analysis of the VIs, band values, and GPP at different
temporal resolutions. This approach provided insights into the temporal dynamics
and patterns of the variables, enabling a more thorough understanding of the
processes and relationships under investigation.

For each site and across all time scales (daily, weekly, and monthly), we remove
GPP values that were less than 1 gC m^-2^ d^-1^ . By doing this, the study aimed
to exclude values that were potentially unreliable or insignificant in terms of
representing meaningful vegetation productivity. GPP values below this threshold
were considered to be either below the detection limit or not enough to
contribute significantly to the overall analysis. Additionally, the removal of
numerous zero GPP values was carried out to prevent undue exaggeration of
regression goodness-of-fit metrics.

The final number of observations per site after selecting the high quality
pixels and GPP observations are shown in @fig-obs_used_analysis

```{r}
#| label: fig-quality_pixels
#| fig-cap: "Total number of observations (pixels) from MODIS classified as high quality (used in the analysis) or other quality (filtered out from the analysis) per site."
#| echo: false
#| message: false
#| warning: false
source("scripts/quality_observations.R")

all %>% 
  ggplot(aes(x = site, fill = quality)) +
  geom_bar(position = "stack") +
  scale_fill_viridis_d(begin = 0.34, end = 0.8) +
  labs(x = "Site", 
       y = "Total observations (pixels)",
       fill = "Quality") +
  theme_bw(base_size = 12)
```

```{r}
#| label: fig-obs_used_analysis
#| fig-cap: "Monthly distribution of high-quality MODIS observations after joining with flux observations containing Gross Primary Productivity (GPP) values higher than 1."
#| echo: false
#| message: false
#| warning: false

# Dataset used to analyze the data
daily_plot_500 %>% 
  # Dataset have all indices per row, so for the same date there are 5 obs
  filter(index == "ndvi_mean") %>% 
  group_by(site, date) %>% 
  tally() %>% 
  group_by(site) %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>% 
  ggplot(aes(x = year_mon, fill = site)) +
  geom_bar(position = "stack") +
  scale_fill_viridis_d(begin = 0.2, end = 0.8) +
  labs(x = "Date",
       y = "Total observations",
       fill = "Site") +
  theme_bw(base_size = 12) 

# daily_plot_500 %>% 
#   pivot_wider(names_from = index, values_from = value) %>% 
#   select(date, total_obs, site) %>% 
#   mutate(year_mon = zoo::as.yearmon(date)) %>% 
#   group_by(site, year_mon) %>% 
#   tally() %>% 
#   ungroup() %>% 
#   mutate(year_mon = as.factor(year_mon)) %>%
#   ggplot(aes(x = year_mon, y = n, fill = site)) +
#   geom_bar(stat = "identity", position = "stack") +
#   scale_fill_viridis_d(begin = 0.2, end = 0.8) +
#   labs(x = "Date",
#        y = "Total observations",
#        fill = "Site") +
#   theme_bw(base_size = 12) +
#   theme(axis.text.x = element_text(angle = 90, h = 1))
```

### Data Analysis

Two distinct modelling approaches were employed to estimate GPP by utilizing the
calculated VIs: a linear model and a Generalized Additive Model (GAM). These
models were created to evaluate the performance of individual indices as well as
their collective performance, while simultaneously considering the effects of
individual site characteristics and the aggregation of sites across various
temporal scales.

This facilitated a comparative analysis of how each individual index
independently explains the variation in GPP per site. Additionally, with the
uniform nature of the three examined forest sites, an evaluation was performed
to explore the GPP relationship without the distinction of the specific sites,
considering each vegetation index individually. Another step of the analysis was
the examination of the relationship between GPP and all indices functioning as
covariates, both on an individual site basis and without site distinction.

We applied the same methodology to the GAM models which allows a better fit for
those cases where the distribution and variability observed in the data is
greater, due to the varying temporal scales. We aimed to evaluate if this
variance could be better explained by this type of model, that might not be
optimal to capture with a linear relationship, potentially leading to greater
residuals.

In total, per each site plus the category of all-sites we created 5 linear
models and 5 GAM models for every timescale (daily, weekly, and monthly): one
per each index (NDVI, CCI, EVI, NIRv) and one with all the indices as covariates
(NDVI + CCI + EVI + NIRv), resulting in a total of 120 models. To compare the
performance of the models we calculated the coefficient of determination (R2),
the Root Mean Squared Error (RMSE), and the Mean Absolute Error (MAE)

\newpage

## Results

### Analysis of GPP-Vegetation Index Relationships Using Linear Models

<!-- ### Monthly GPP and VIs relations -->

The @tbl-lm_monthly_results provides a summary of linear models used for GPP
estimation at each site, employing the vegetation indices as predictors. For
each site and predictor, the table includes the relevant model summary
statistics, such as R2, MAE, and RMSE. More metrics such as the p-value,
adjusted r-squared, the Akaike Information Criterion (AIC), and Bayesian
Information Criterion (BIC) are displayed in the
@tbl-complete_lm_monthly_results for monthly results,
@tbl-complete_lm_weekly_results for weekly results, and the
@tbl-complete_lm_daily_results for daily outputs. A residuals distribution for
each of the models is in @fig-lm_residuals

Our findings shows that when using all the indices as covariates within the
linear model, the model's performance demonstrates better outcomes compared to
using any single index alone across all scenarios. In the assessment of
individual VI performance on a monthly basis, CCI tend to perform better than
EVI, NDVI, and NIRv. However it is important to note that for the Bartlett and
Michigan sites, while CCI shows favourable predictive accuracy, its advantage
over the NIRv and EVI is slightly better and mostly due to the MAE and RMSE
results.

The NDVI displays relatively diminished performance, indicating a 9% and 15%
reduction in predictive accuracy compared to CCI in the Bartlett and Michigan
sites, respectively. In contrast, EVI exhibits less favourable predictive
results for the Borden site and the aggregated sites category, although in the
specific context of the Borden site, its performance aligns closely with NDVI
and NIRv. Among the individual sites, Bartlett stands out with the most
favourable predictive outcomes, while the aggregated sites category yield the
least favourable predictive results.

In the case of the weekly models, NDVI records the least favourable results when
assessed at the three individual sites, while EVI demonstrated its weakest
predictive capabilities when all sites were treated as a single entity. However,
for this case, EVI shows marginal differences with respect to individual indices
such as NDVI and NIRv (1% variance explained and no more than 0.02 in RMSE). The
best performant individual indices were CCI for Borden and the aggregated sites,
NIRv for Michigan and EVI for Bartlett. Nonetheless, those superior performances
are subtle when compared with the other individual indices. On a weekly basis,
differences in variability explained between the best and least performing
models range from 11% to 20%. Bartlett and Michigan sites consistently yield the
most accurate predictive models.

On a daily basis, CCI outperforms the rest of the individual indices in 3 cases:
for Bartlett by 2% in variance explanation compared with EVI, for Borden by 1%
compared with EVI, and for the combined sites dataset by 10%. Nonetheless is
worth to mention that the variance explained in any of the models by individual
indices in Borden or the combined site is less than 50% in most of the cases and
the error scores being the highest. In the case of Michigan, the best performing
individual index was EVI which outperformed the next best-performing individual
index NIRv by 3%. Generally, the Bartlett and Michigan sites consistently
yielded the most accurate predictive models across various configurations.
Conversely, the Borden site consistently exhibited the poorest model performance
across all scenarios.

Overall, when comparing individual indices, CCI consistently performed better
across different timeframes in terms of variance explainability and error
metrics. Nonetheless, those differences are subtle compared to EVI and NIRv. In
contrast, NDVI consistently performs less favorably across all evaluated
timeframes. Notably, models based on monthly values consistently exhibit better
performance than those based on weekly or daily values as ilustrated in
@fig-lm_metrics.

```{=tex}
\begingroup
\setlength{\LTleft}{0pt minus 500pt}
\setlength{\LTright}{0pt minus 500pt}
\fontsize{5pt}{7pt}\selectfont
\addtolength{\tabcolsep}{-3pt}
```
```{r lm_monthly}
#| label: tbl-lm_monthly_results
#| tbl-cap: "Summary of Linear models for GPP estimation using the vegetation indices on a monthly (a), weekly (b), and daily (c) basis.  MAE and RMSE metrics units are gC m-⁻² d-⁻¹"
#| tbl-colwidths: [50,50]
#| echo: false
#| message: false
#| warning: false
source("scripts/lm_preparation.R")

create_metrics_table <- function(data) {
  data %>% 
    gt() %>%
    tab_spanner(label = md("NDVI"), columns = ends_with("NDVI")) %>%
    tab_spanner(label = "EVI", columns = ends_with("EVI")) %>%
    tab_spanner(label = "NIRv", columns = ends_with("NIRv")) %>%
    tab_spanner(label = "CCI", columns = ends_with("CCI")) %>% 
    tab_spanner(label = "All", columns = ends_with("All")) %>% 
    cols_label(
      .list = list(
        "site" = "Site",
        "r.squared_EVI" = "R2",
        "mae_EVI" = "MAE",
        "rmse_EVI" = "RMSE",
        "r.squared_NDVI" = "R2",
        "mae_NDVI" = "MAE",
        "rmse_NDVI" = "RMSE",
        "r.squared_NIRv" = "R2",
        "mae_NIRv" = "MAE",
        "rmse_NIRv" = "RMSE",
        "r.squared_CCI" = "R2",
        "mae_CCI" = "MAE",
        "rmse_CCI" = "RMSE",
        "r.squared_All" = "R2",
        "mae_All" = "MAE",
        "rmse_All" = "RMSE"
      )
    ) %>% 
    cols_align(
      align = "center",
      columns = 2:16
    ) %>% 
    fmt_number(
      columns = 2:16,
      decimals = 2) %>% 
    tab_options(
      row_group.background.color = "#E9E0E1",
      row_group.font.weight = "bold"
    ) %>% 
    cols_width(everything() ~ px(50))
}

# Monthly table
vis_site_glance_monthly %>% 
  bind_rows(all_sites_glance_monthly,
            all_sites_all_vis_glance_monthly,
            all_vis_glance_monthly) %>% 
  select(site, index, r.squared, mae, rmse) %>%
  pivot_wider(names_from = index, 
              values_from = c(r.squared, mae, rmse)) %>%
  create_metrics_table()


# Weekly table
vis_site_glance_weekly  %>%
  bind_rows(all_sites_glance_weekly,
            all_sites_all_vis_glance_weekly,
            all_vis_glance_weekly) %>% 
  select(site, index, r.squared, mae, rmse) %>%
  pivot_wider(names_from = index, 
              values_from = c(r.squared, mae, rmse)) %>%
  create_metrics_table()

# Daily table
vis_site_glance_daily %>%
  bind_rows(all_sites_glance_daily,
            all_sites_all_vis_glance_daily,
            all_vis_glance_daily)  %>% 
  select(site, index, r.squared, mae, rmse) %>%
  pivot_wider(names_from = index, 
              values_from = c(r.squared, mae, rmse)) %>%
  create_metrics_table()
```

```{=tex}
\endgroup
```
```{r lm_barplot_metrics}
#| label: fig-lm_metrics
#| fig-cap: "Summary of Linear models for GPP estimation using the vegetation indices on a monthly (a), weekly (b), and daily (c) basis. MAE and RMSE metrics units are gC m⁻² d⁻¹"
#| fig-width: 7
#| fig-height: 9
#| echo: false
#| message: false
#| warning: false
library(cowplot)

# Create a function to generate the plot
create_metrics_plot <- function(timescale, y_var, ylim_range) {
  get(paste0("vis_site_glance_", timescale)) %>% 
    select(site, index, {{ y_var }}) %>%
    bind_rows(get(paste0("all_sites_glance_", timescale)),
               get(paste0("all_sites_all_vis_glance_", timescale)),
              get(paste0("all_vis_glance_", timescale))) %>% 
    ggplot(aes(x = site, y = .data[[y_var]], fill = index)) +
    geom_col(position = "dodge") +
    coord_cartesian(ylim = ylim_range) +
    scale_fill_viridis_d() +
    labs(x = "", fill = "Index") +
    theme_minimal_hgrid(font_size = 12) +
    theme(axis.text.x = element_text(angle = 45, h = 1))
}

# Response variables are going to be the same:
response_vars <- c("r.squared", "mae", "rmse")

# Monthly plots
monthly_metrics_plots <- map2(response_vars, 
     list(c(0.3, 1), c(0.5, 4), c(0.5, 5)), 
     ~ create_metrics_plot("monthly", .x, .y))

# Weekly plots
weekly_metrics_plots <- map2(response_vars, 
              list(c(0.3, 1), c(0.5, 4), c(0.5, 5)),
              ~ create_metrics_plot("weekly", .x, .y))

# Daily plots
daily_metrics_plots <- map2(response_vars, 
              list(c(0.3, 1), c(0.5, 4), c(0.5, 5)),
              ~ create_metrics_plot("daily", .x, .y))

# Grid the plots as should go in the chapter
lm_metrics_plots <- plot_grid(
  monthly_metrics_plots[[1]] + labs(y = expression(R^{"2"})) + theme(legend.position = "none"),
  weekly_metrics_plots[[1]] +  labs(y = expression(R^{"2"})) + theme(legend.position = "none"),
  daily_metrics_plots[[1]] +   labs(y = expression(R^{"2"})) + theme(legend.position = "none"),
  monthly_metrics_plots[[2]] + labs(y = "MAE") + theme(legend.position = "none"),
  weekly_metrics_plots[[2]] +  labs(y = "MAE") + theme(legend.position = "none"),
  daily_metrics_plots[[2]] +   labs(y = "MAE") + theme(legend.position = "none"),
  monthly_metrics_plots[[3]] + labs(y = "RMSE") + theme(legend.position = "none"),
  weekly_metrics_plots[[3]] +  labs(y = "RMSE") + theme(legend.position = "none"),
  daily_metrics_plots[[3]] +   labs(y = "RMSE") + theme(legend.position = "none"),
  nrow = 3,
  ncol = 3,
  labels = c("A", "B", "C"))

plot_legend <- get_legend(
  monthly_metrics_plots[[1]] + 
    guides(color = guide_legend(nrow = 3)) 
)

plot_grid(lm_metrics_plots, plot_legend, ncol = 2, rel_widths = c(1, .1))
rm(plot_legend)
```

### Analysis of GPP-Vegetation Index Relationships Using GAM Models

In @tbl-gam_model_results, we present a summary of the results obtained from the
GAM models used for GPP estimation at each site, employing the vegetation
indices as predictors. To compare between the models, we include in the table
relevant model summary statistics, such as R2, MAE, and RMSE. Furthermore,
additional metrics such as the p-value, the F statistic (f), effective degrees
of freedom (edf), and the Akaike Information Criterion (AIC), are displayed in
the @tbl-gam_daily_model_results_complete_a, and
@tbl-gam_daily_model_results_complete_b for daily models,
@tbl-gam_weekly_model_results_complete_a, and
@tbl-gam_weekly_model_results_complete_b for weekly outputs, and for monthly
results the @tbl-gam_monthly_model_results_complete_a, and
@tbl-gam_monthly_model_results_complete_b. A residuals distribution for each of
the models is in @fig-gam_residuals

<!-- explains more x% variance than -->

When using GAM models at a monthly basis, no single VI demonstrates consistent
superiority over the others. For the all sites category, CCI have a 3% better
variance explanation than NDVI, which is the second best model with a 75%. In
the case of Michigan EVI and NIRv were the best individual indices with a 96% of
the variance in GPP, but EVI had slightly lower error metrics values in MAE
(0.01) and RMSE (0.02). Bartlett had a better model performance when using CCI
with a 91% of variance explained with the lowest error metrics among all the GAM
monthly models. It's important to highlight that for Michigan and Bartlett,
implementing a GAM model using all the VIs as covariates posed challenges due to
limited observations for the model parameters, raising concerns of potential
overfitting.

Finally, it's noteworthy that NDVI on a monthly basis displayed suboptimal
performance in two of the sites (Michigan and Bartlett) with a difference of 10%
and 13% with the best performing models respectively, while EVI exhibited less
favourable results in just the Borden site.

On a weekly basis, models incorporating all VIs as covariates consistently
obtained better performance compared to any individual VI, irrespective of the
site. Specifically when evaluating the all sites category, the inclusion of all
indices as covariates yielded a 7% increase in variance explanation compared
with the best individual VI result CCI. Further, for Michigan, this improvement
amounted to 2% compared to EVI, 1% for Bartlett in contrast to EVI, and a 2%
enhancement for Borden when compared with EVI.

Conversely, NDVI showed a diminished performance as an individual index when
compared with all the other individual VIs. In the context of all sites, NDVI
yielded 7% less variance explanation than CCI. Notably, for Michigan, NDVI's
performance lagged by 17% compared to EVI, Bartlett an 18% reduction relative to
EVI, and Borden exhibited a 9% deficit when contrasted with EVI.

On a daily basis, when considering the all sites category, the model with all
the VIs as covariates explained 4% more variance in GPP when compared with the
best performing individual VI CCI. In the case of Bartlett, the increased was
also 4% but in this case the best individual performing VI was EVI. For
Michigan, the model using all VIs as covariates outperformed EVI by 5% in GPP
prediction, and for Borden it was a 9% improvement when compared with NIRv.

Among the individual VIs, NDVI consistently demonstrated the poorest performance
across al four cases. As an individual VI, EVI performed better in Bartlett,
Michigan, and Borden. However is worth noting that for Borden the variance
explained was limited to 50%.

In summary, among the three individual sites, Bartlett consistently produced the
most favourable results in terms of models explaining variance and yielding
lower residuals, followed by Michigan. In the case of Borden, when employing
individual indices, the models struggled to achieve a variance explanation
exceeding 50%.

<!-- ### Daily and weekly GPP and VIs relations -->

<!-- **TODO: Single vs Individual GAM model** -->

<!-- *This is a section under construction. Several models are presented in order -->

<!-- to discuss which one will be better to describe the patterns found* -->

<!-- Richard's suggestion is to create a single model using all indices as covariates. -->

<!-- The notes are described also in the issue [Ref #49](https://github.com/ronnyhdez/thesis_msc/issues/49) -->

<!-- **Single model** -->

<!--  - Using all the indices that I have as multiple predictors will allow me to  -->

<!--  explore the relation between GPP and each VI while also considering potential -->

<!--  interactions. But this will be an examination of the combined effects of the  -->

<!--  VIs on GPP and determine their relative importance. -->

<!-- **Individual model** -->

<!--  - A separate GAM model for each VI will allow me to explore the relation  -->

<!--  between GPP and each VI separately, without the exploration of potential  -->

<!--  interactions between the VIs.  -->

<!--  - But, do I want to explore if there are interactions? If there is an  -->

<!--  interaction, what would be the meaning of an interaction between indices? -->

<!--  - With the individual GAM models, I think that I can uncover individual trends, that otherwise would be hidden among all the VIs. [**simpson's paradox**](https://en.wikipedia.org/wiki/Simpson%27s_paradox) -->

```{=tex}
\begingroup
\setlength{\LTleft}{0pt minus 500pt}
\setlength{\LTright}{0pt minus 500pt}
\fontsize{5pt}{7pt}\selectfont
\addtolength{\tabcolsep}{-3pt}
```
```{r}
#| label: tbl-gam_model_results
#| tbl-cap: "Summary of GAM models for GPP estimation using the vegetation indices on a monthly (a), weekly (b), and daily (c) basis.  MAE and RMSE metrics units are gC m-⁻² d-⁻¹" 
#| echo: false
#| message: false
#| warning: false
source("scripts/gam_preparation.R")

create_metrics_table <- function(data) {
  data %>% 
    pivot_wider(names_from = index,
                values_from = c(rsq, mae, rmse)) %>% 
    gt() %>%
    tab_spanner(label = md("NDVI"), columns = contains("NDVI")) %>%
    tab_spanner(label = "EVI", columns = contains("EVI")) %>%
    tab_spanner(label = "NIRv", columns = contains("NIRv")) %>%
    tab_spanner(label = "CCI", columns = contains("CCI")) %>% 
    tab_spanner(label = "All", columns = contains("All")) %>% 
    cols_label(
      .list = list(
        "site" = "Site",
        "rsq_evi_mean" = "R2",
        "mae_evi_mean" = "MAE",
        "rmse_evi_mean" = "RMSE",
        "rsq_ndvi_mean" = "R2",
        "mae_ndvi_mean" = "MAE",
        "rmse_ndvi_mean" = "RMSE",
        "rsq_nirv_mean" = "R2",
        "mae_nirv_mean" = "MAE",
        "rmse_nirv_mean" = "RMSE",
        "rsq_cci_mean" = "R2",
        "mae_cci_mean" = "MAE",
        "rmse_cci_mean" = "RMSE",
        "rsq_All" = "R2",
        "mae_All" = "MAE",
        "rmse_All" = "RMSE"
      )
    ) %>% 
    cols_align(
      align = "center",
      columns = 2:16
    ) %>% 
    fmt_number(
      columns = 2:16,
      decimals = 2) %>% 
    tab_options(
      row_group.background.color = "#E9E0E1",
      row_group.font.weight = "bold"
    ) %>% 
    cols_width(everything() ~ px(50)) 
}

# Monthly table
all_sites_gam_monthly %>% 
  bind_rows(vis_sites_gam_monthly,
            all_sites_all_vis_gam_monthly,
            all_vis_gam_monthly,
  ) %>% 
  create_metrics_table()

# Weekly table
all_sites_gam_weekly %>% 
  bind_rows(vis_sites_gam_weekly,
            all_sites_all_vis_gam_weekly,
            all_vis_gam_weekly,
  ) %>% 
  create_metrics_table()

# Daily table
all_sites_gam_daily %>% 
  bind_rows(vis_sites_gam_daily,
            all_sites_all_vis_gam_daily,
            all_vis_gam_daily,
  ) %>% 
  create_metrics_table()
```

```{=tex}
\endgroup
```
```{r gam_barplot_metrics}
#| label: fig-gam_metrics
#| fig-cap: "Summary of GAM models for GPP (gC m-² d-¹) estimation using the vegetation indices. Column A represents the metrics for the monthly models, B the weekly, and C the daily metrics. MAE and RMSE metrics units are gC m⁻² d⁻¹"
#| fig-width: 7
#| fig-height: 9
#| echo: false
#| message: false
#| warning: false

# Create a function to generate the plot
create_metrics_plot <- function(timescale, y_var, ylim_range) {
  get(paste0("all_sites_gam_", timescale)) %>% 
    # select(site, index, {{ y_var }}) %>%
    bind_rows(get(paste0("vis_sites_gam_", timescale)),
              get(paste0("all_sites_all_vis_gam_", timescale)),
              get(paste0("all_vis_gam_", timescale))) %>% 
    mutate(index = case_when(
      index == "evi_mean" ~ "EVI",
      index == "ndvi_mean" ~ "NDVI",
      index == "nirv_mean" ~ "NIRv",
      index == "cci_mean" ~ "CCI",
      .default = index
    )) %>% 
    ggplot(aes(x = site, y = .data[[y_var]], fill = index)) +
    geom_col(position = "dodge") +
    coord_cartesian(ylim = ylim_range) +
    scale_fill_viridis_d() +
    labs(x = "", fill = "Index") +
    theme_minimal_hgrid(font_size = 12) +
    theme(axis.text.x = element_text(angle = 45, h = 1))
}

# Response variables are going to be the same:
response_vars <- c("rsq", "mae", "rmse")

# Monthly plots
monthly_metrics_plots <- map2(response_vars, 
     list(c(0.3, 1), c(0.5, 4), c(0.8, 5)), 
     ~ create_metrics_plot("monthly", .x, .y))

# Weekly plots
weekly_metrics_plots <- map2(response_vars, 
              list(c(0.3, 1), c(0.5, 4), c(0.8, 5)),
              ~ create_metrics_plot("weekly", .x, .y))

# Daily plots
daily_metrics_plots <- map2(response_vars, 
              list(c(0.3, 1), c(0.5, 4), c(0.8, 5)),
              ~ create_metrics_plot("daily", .x, .y))

# Grid the plots as should go in the chapter
gam_metrics_plots <- plot_grid(
  monthly_metrics_plots[[1]] + labs(y = expression(R^{"2"})) + theme(legend.position = "none"),
  weekly_metrics_plots[[1]] +  labs(y = expression(R^{"2"})) + theme(legend.position = "none"),
  daily_metrics_plots[[1]] +   labs(y = expression(R^{"2"})) + theme(legend.position = "none"),
  monthly_metrics_plots[[2]] + labs(y = "MAE") + theme(legend.position = "none"),
  weekly_metrics_plots[[2]] +  labs(y = "MAE") + theme(legend.position = "none"),
  daily_metrics_plots[[2]] +   labs(y = "MAE") + theme(legend.position = "none"),
  monthly_metrics_plots[[3]] + labs(y = "RMSE") + theme(legend.position = "none"),
  weekly_metrics_plots[[3]] +  labs(y = "RMSE") + theme(legend.position = "none"),
  daily_metrics_plots[[3]] +   labs(y = "RMSE") + theme(legend.position = "none"),
  nrow = 3,
  ncol = 3,
  labels = c("A", "B", "C"))

plot_legend <- get_legend(
  monthly_metrics_plots[[1]] + 
    guides(color = guide_legend(nrow = 3)) 
)

plot_grid(gam_metrics_plots, plot_legend, ncol = 2, rel_widths = c(1, .1))
rm(plot_legend)
```

### LM vs GAM

Monthly GAM applied in the context of Michigan exhibited a
better performance when employing EVI and NIRv in comparison to LMs with all
vegetation indices as covariates. Regarding weekly GAM models, they demonstrated
improved performance compared to LM across all sites, except for Borden, where
both models yielded equivalent R² values. Notably, despite similar R² values,
the GAM model consistently displayed lower error metrics, suggesting its
capacity to better accommodate potential nonlinear relationships and produce
more accurate predictions. This adaptability was particularly evident when
addressing the daily, weekly, or monthly variations in GPP. Furthermore, it is
crucial to highlight that monthly models, both in LM and GAM frameworks,
consistently exhibited superior metrics compared to their weekly and daily
counterparts, as illustrated in @fig-gam_metrics and @fig-lm_metrics.

```{r gpp_vi_realtions_all_sites}
#| label: fig-gpp_vi_relation_daily
#| fig-cap: "Scatterplot of MODIS 500m derived VIs and GPP with daily values. Every observation corresponds to the observed GPP from a flux tower site. Total observations corresponds to the number of observations used to obtain the mean of the vegetation index (NDVI, NIRv, CCI and EVI). The red line indicates the GAM fit."
#| fig-width: 7
#| fig-height: 5
#| echo: false
#| message: false
#| warning: false

## Make sure the source of file "scripts/trend_plots.R" was successful
etiquetas <- c(
  "evi_mean" = "EVI",
  "ndvi_mean" = "NDVI",
  "nirv_mean" = "NIRv",
  "kndvi_mean" = "kNDVI",
  "cci_mean" = "CCI",
  "Borden" = "Borden",
  "Michigan" = "Michigan",
  "Bartlett" = "Bartlett"
)

# daily_grid <-
  daily_plot_500 %>% 
  filter(index != "kndvi_mean") %>% 
  # filter(ndvi_mean > 3 & gpp_dt_vut_ref < 20)
  mutate(month = lubridate::month(date, label = TRUE)) %>% 
  ggplot(aes(x = value, y = gpp_dt_vut_ref)) +
  geom_point(aes(color = month, size = total_obs), alpha = 0.8) +
  geom_smooth(method = "gam", se = FALSE, color = "#E20D6A") +
  scale_color_viridis_d() + 
  scale_fill_viridis_d(aes(month)) +
  facet_grid(site~index, scales = "free", labeller = as_labeller(etiquetas)) +
  scale_size(range = c(.1, 3), name = "Total observations") +
  scale_x_continuous(breaks = seq(-1, 1, by = .1)) +
  scale_y_continuous(breaks = seq(0, 38, by = 4)) +
  labs(x = "Index value", 
       y = expression(GPP~(gC~m^{"-2"}~d^-1)),
       color = "Month") +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(angle = 60, h = 1)) +
  guides(size = guide_legend(order = 1, keyheight = 0.1),
         col = guide_legend(order = 2, keyheight = 0.1)) +
  theme(strip.text = element_text(size = 10, color = "black"),
        strip.background = element_blank())
```

```{r gpp_vi_realtions_all_sites}
#| label: fig-gpp_vi_relation_weekly
#| fig-cap: "Scatterplot of MODIS 500m derived VIs and GPP with weekly values. Every observation corresponds to the observed GPP from a flux tower site. Total observations corresponds to the number of observations used to obtain the mean of the vegetation index (NDVI, NIRv, CCI and EVI). The red line indicates the GAM fit."
#| fig-width: 7
#| fig-height: 5
#| echo: false
#| message: false
#| warning: false
# weekly_grid <-
  weekly_plot_500 %>% 
  filter(index != "kndvi_mean") %>% 
  mutate(month = lubridate::month(date_start, label = TRUE)) %>% 
  ggplot(aes(x = value, y = gpp_dt_vut_ref)) +
  geom_point(aes(color = month, size = total_obs), alpha = 0.8) +
  geom_smooth(method = "gam", se = FALSE, color = "#E20D6A") +
  scale_color_viridis_d() + 
  scale_fill_viridis_d(aes(month)) +
  facet_grid(site~index, scales = "free", labeller = as_labeller(etiquetas)) +
  scale_size(range = c(.1, 3), name = "Total observations") +
  scale_x_continuous(breaks = seq(-1, 1, by = .1)) +
  scale_y_continuous(breaks = seq(0, 38, by = 4)) +
  labs(x = "Index value", 
       y = expression(GPP~(gC~m^{"-2"}~d^-1)),
       color = "Month") +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(angle = 60, h = 1)) +
  guides(size = guide_legend(order = 1, keyheight = 0.1),
         col = guide_legend(order = 2, keyheight = 0.1)) +
  theme(strip.text = element_text(size = 10, color = "black"),
        strip.background = element_blank())
```

```{r gpp_vi_realtions_all_sites}
#| label: fig-gpp_vi_relation_monthly
#| fig-cap: "Scatterplot of MODIS 500m derived VIs and GPP with monthly values. Every observation corresponds to the observed GPP from a flux tower site. Total observations corresponds to the number of observations used to obtain the mean of the vegetation index (NDVI, NIRv, CCI and EVI). The red line indicates the GAM fit."
#| fig-width: 7
#| fig-height: 5
#| echo: false
#| message: false
#| warning: false

# monthly_grid <-
  monthly_plot_500 %>% 
  filter(index != "kndvi_mean") %>% 
  mutate(month = lubridate::month(date, label = TRUE)) %>% 
  ggplot(aes(x = value, y = gpp_dt_vut_ref)) +
  geom_point(aes(color = month, size = total_obs), alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "#E20D6A") +
  scale_color_viridis_d() + 
  scale_fill_viridis_d(aes(month)) +
  facet_grid(site~index, scales = "free", labeller = as_labeller(etiquetas)) +
  scale_size(range = c(.1, 3), name = "Total observations") +
  scale_x_continuous(breaks = seq(-1, 1, by = .1)) +
  scale_y_continuous(breaks = seq(0, 18, by = 4)) +
  labs(x = "Index value", 
       y = expression(GPP~(gC~m^{"-2"}~d^-1)),
       color = "Month") +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(angle = 60, h = 1)) +
  guides(size = guide_legend(order = 1, keyheight = 0.1),
         col = guide_legend(order = 2, keyheight = 0.1)) +
  theme(strip.text = element_text(size = 10, color = "black"),
        strip.background = element_blank())

# plot_grid(daily_grid,
#           weekly_grid,
#           monthly_grid,
#           nrow = 3,
#           labels = c('A', 'B', 'C'))
```

\newpage

## Discussion

Based on the RMSE, NDVI performed less favourably on two out of the three sites
compared with the other indices. However, the performance difference among the
sites was not substantial, with the Bartlett experimental forest exhibiting
slightly better results on all of their GPP \~ VIs relations compared to the
other sites. These findings suggest that while NDVI may not be the optimal
vegetation index for GPP estimation at these sites, there may still be
variations in performance across different locations.

\newpage

## References

**Just the references of this chapter, that also needs to be included in the
final references.**

::: {#refs}
:::
